---
title: "A short introduction to SHIP"
author: "SHIP authors"
format:
  html:
    toc: true
    code-fold: true
    html-math-method: mathjax
execute:
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{A short introduction to SHIP}
  %\VignetteEngine{quarto::html}
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(SHIP)
```

SHIP estimates covariance matrices by combining the sample covariance matrix
with a structured target. The final estimate has the form

$$
\widehat{\Sigma} = \lambda T + (1 - \lambda)S,
$$

where $S$ is the sample covariance matrix, $T$ is a target, and
$0 \leq \lambda \leq 1$ controls the amount of shrinkage.

## The example data

The package includes `expl`, a small microarray data set with 102 observations
and 100 genes. The accompanying `genegroups` list contains KEGG pathway IDs
when they are available; `NA` means that no pathway information is available
for that gene.

```{r}
#| label: data-summary

 data("expl")
dim(expl$x)
length(expl$genegroups)
sum(vapply(expl$genegroups, function(group) !all(is.na(group)), logical(1)))
```

## Comparing covariance targets

`build.target()` constructs several targets. `D` and `F` use only the data;
`G`, `Gpos`, `Gstar`, and `cor` also use the pathway groups.

```{r}
#| label: target-summary

target_types <- c("D", "F", "G", "Gpos", "Gstar", "cor")
targets <- setNames(
  lapply(target_types, function(type) {
    build.target(expl$x, expl$genegroups, type = type)
  }),
  target_types
)

target_summary <- data.frame(
  target = names(targets),
  dimension = vapply(targets, function(target) paste(dim(target), collapse = " x "), character(1)),
  nonzero = vapply(targets, function(target) sum(target != 0), integer(1)),
  row.names = NULL
)
target_summary
```

The diagonal target keeps the sample variances and sets covariances to zero.
The constant-correlation target uses one average correlation for all
off-diagonal entries. The knowledge-based targets restrict correlations to
genes that share a pathway; `Gpos` keeps positive links, while `Gstar`
separates positive and negative links.

## Estimating a covariance matrix

The target can then be passed to `shrink.estim()`.

```{r}
#| label: shrinkage

estimate <- shrink.estim(expl$x, targets[["G"]])

c(
  lambda = estimate$lambda,
  sample_variance = sum(diag(cov(expl$x))),
  estimated_variance = sum(diag(estimate$shrink.cov))
)
```

A value of `lambda` close to zero keeps more of the empirical covariance
matrix. A value close to one gives more weight to the structured target.

The target definitions are based on Schaefer and Strimmer (2005) and
Jelizarow et al. (2010), *Over-optimism in bioinformatics: an illustration*.
